---
output:
  pdf_document: default
  html_document: default
---
# Faulty LED seven-edge digit recognition

#### Josep Fortiana 2019-11-26

## Original problem

The problem involves representations of digits by seven LED lights which are selectively lit to form each digit. It is described in the book: Leo Breiman, Jerome Friedman, R.A. Olshen and Charles J. Stone (1984) _Classification and Regression Trees_ (BFOS), p. 43-49.

`Digits.plot.r` contains auxiliary functions for visualizing them.

```{r}
source("Digits.plot.r")
```

The figure below shows the numbering of the seven edges used to form the digits

```{r}
options(repr.plot.width=4,repr.plot.height=4)
plot.digits.matrix.with.segment.indexes()
```

Here we see the ten digits, represented by the seven edges.

```{r}
options(repr.plot.width=7.0,repr.plot.height=2.5)
show.all.digits()
```

The file `digits.csv` contains $500$ sets (rows), each with seven ones and zeros, representing `ON`/`OFF` status for each of the seven edges, numbered as above, plus a label, the intended digit. Most rows, however, contain faulty edges, set `ON` when they should be `OFF` and vice-versa. For instance, the following picture shows a faulty '5' where the sixth edge failed to light on.

```{r}
options(repr.plot.width=2.5,repr.plot.height=3.0)
v<-as.logical(c(1,1,0,1,0,0,1))
plot.faulty.digit(v)
```

Each row in `digits.csv` contains an additional variable, `DIGIT`, with the name of the intended digit.

```{r}
digits.data<-read.csv("digits.csv")
str(digits.data)
```

```{r}
levels(digits.data$digit)
```

Better with shorter (and equal length) names for levels

```{r}
#install.packages("dplyr",dependencies=TRUE,repos="https://cloud.r-project.org")
require(dplyr)
```

```{r}
digits.data$digit<-recode(digits.data$digit,"one"="1","two"="2",
          "three"="3","four"="4","five"="5","six"="6","seven"="7",
          "eight"="8","nine"="9","zero"="0")
```

```{r}
str(digits.data)
```

```{r}
table(digits.data$digit)
```

## Generating more data

Possibly the `digits.csv` dataset is extracted from the collection of datasets in the _Splus_ program. In the BFOS book they mention an (unpublished, I believe) training dataset with $200$ observations and a test sample of $5000$. The generator program by David W. Aha, in C, can be found at the [UCI Machine Learning repository](https://archive.ics.uci.edu/ml/datasets/LED+Display+Domain).

A better alternative is the `rdigitsBFOS()` function from the `gencve` package, which generates digits with a given flipping probability `alpha` (that is, the probability of one of the seven LEDs changing its correct status to the opposite one). As a bonus the function computes the Bayes optimal missclassification rate, which can be used to have an objective lower
bound for the quality of any classification method.

```{r}
#install.packages("gencve",dependencies=TRUE,repos="https://cloud.r-project.org")
require(gencve)
```